sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-03 doc/paper-jss/inla-comparison.Rmd
figure captions for inla appendix

Comparing the fixed effects shows the estimates from all three models are also very similar.

```{r coef-plot, fig.asp=0.4, out.width="85%", echo=FALSE}
inla_coefs <- m_inla$summary.fixed[, c("0.025quant", "mean", "0.975quant")] %>%
  rename(estimate = mean, conf.low = `0.025quant`, conf.high = `0.975quant`) %>%
  mutate(term = unique(sdmTMB_coefs$term), model = "INLA")

@@

We can then compare the distribution of these \pkg{sdmTMB} simulations to the marginal posterior densities from \pkg{INLA}
From \pkg{INLA}, we can extract the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the \pkg{sdmTMB} `gather_sims()` function.
Alternatively, we could use `inla.posterior.sample(result = m_inla, n = 1000)` to sample from the \pkg{INLA} model posterior assuming we had set\
`control.compute` ` = list(config = TRUE)`.

@@

In this plot, the blue histograms are from \pkg{sdmTMB} and the density lines are from \pkg{INLA}:

```{r compare-plot, out.width="6in"}
ggplot(post_inla2) +
  geom_histogram(
    data = post_sdmTMB, aes(.value, after_stat(density)), bins = 30,
    fill = "blue", alpha = 0.2
  ) +
  geom_line(data = post_inla2, aes(x, y), inherit.aes = FALSE) +
  facet_wrap(~variable, scales = "free")
```

@@

We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the \pkg{sdmTMB} package to map occurrence probabilities in each year in space.
For \pkg{sdmTMB}, the resulting values are in link space and require the appropriate inverse-link transformation (always contained in `model$family$linkinv()`, equivalent to inverse logit here), to produce occurrence probabilities:

@@

```{r maps, out.width="5.5in"}
inverse_link <- m$family$linkinv
ggplot(pred) +
  geom_raster(aes(X, Y, fill = inverse_link(est))) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

We can get similar predictions with \pkg{inlabru} using only a few additional steps: we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and the appropriate inverse-link transformation:

```{r inlabru-pred, out.width="5.5in"}
nd_bru <- nd
nd_bru <- dplyr::left_join(nd_bru, yr_lu, by = "year")
sp::coordinates(nd_bru) <- c("X", "Y")
nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))

>>>>

Comparing the fixed effects shows the estimates from all three models are also very similar (Fig.\ \ref{fig:coef-plot}).

```{r coef-plot, fig.asp=0.4, out.width="85%", echo=FALSE, fig.cap="Fixed effect estimates from all three models with 95 percent CI.", fig.pos='h', fig.align='center'}
inla_coefs <- m_inla$summary.fixed[, c("0.025quant", "mean", "0.975quant")] %>%
  rename(estimate = mean, conf.low = `0.025quant`, conf.high = `0.975quant`) %>%
  mutate(term = unique(sdmTMB_coefs$term), model = "INLA")

@@

We can then compare the distribution of these \pkg{sdmTMB} simulations to the marginal posterior densities from \pkg{INLA}
From \pkg{INLA} (Fig.\ \ref{fig:compare-plot}). 
We do this by extracting the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the \pkg{sdmTMB} `gather_sims()` function.
Alternatively, we could use `inla.posterior.sample(result = m_inla, n = 1000)` to sample from the \pkg{INLA} model posterior assuming we had set\
`control.compute` ` = list(config = TRUE)`.

@@

<!-- In this plot, the blue histograms are from \pkg{sdmTMB} and the density lines are from \pkg{INLA}: -->

```{r compare-plot, out.width="6in", fig.cap="Blue histograms represent simulated draws from the joint precision matrix of the sdmTMB model compared with the marginal posterior densities from the INLA model.", fig.align='center'}
ggplot(post_inla2) +
  geom_histogram(
    data = post_sdmTMB, aes(.value, after_stat(density)), bins = 30,
    fill = "blue", alpha = 0.2
  ) +
  geom_line(data = post_inla2, aes(x, y), inherit.aes = FALSE) +
  facet_wrap(~variable, scales = "free")
```

@@

We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the \pkg{sdmTMB} package to map occurrence probabilities in each year in space (Fig.\ \ref{fig:maps}).
For \pkg{sdmTMB}, the resulting values are in link space and require the appropriate inverse-link transformation (always contained in `model$family$linkinv()`, equivalent to inverse logit here), to produce occurrence probabilities:

@@

```{r maps, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the sdmTMB model.", fig.align='center'}
inverse_link <- m$family$linkinv
ggplot(pred) +
  geom_raster(aes(X, Y, fill = inverse_link(est))) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

We can get similar predictions with \pkg{inlabru} using only a few additional steps: we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and the appropriate inverse-link transformation (Fig.\ \ref{fig:inlabru-pred}).

```{r inlabru-pred, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the inlabru model.", fig.align='center'}
nd_bru <- nd
nd_bru <- dplyr::left_join(nd_bru, yr_lu, by = "year")
sp::coordinates(nd_bru) <- c("X", "Y")
nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))
